---
title: "PCA"
output: pdf_document
date: "2023-03-16"
---
# Set the working directory
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dirname(dirname(rstudioapi::getActiveDocumentContext()$path)))

library(dplyr)
```

## Read Data

#```{r, echo=FALSE}
# Go to the MD project folder before running the script
#dir <- "C:/Users/polcg/repos/MD" # for some reason the it can't access current directory
# Import preprocessed data
# db <- read.csv(paste0(dir,"/data/Preprocessed_data.csv"),header=T, sep=",", strip.white = TRUE) # take out trailing whitespaces

# read dataframe back in from RDS file
#db <- readRDS(paste0(dir,"/data/Preprocessed_data_rds.rds"))
#```

# Try this to read the data
```{r, echo=FALSE}
# Get the data
db <- readRDS("./data/Preprocessed_data_rds.rds")
```


## We check for missing data

```{r, echo=FALSE}
# Check if there are any missing values, there should not be because they have been imputed, or removed
sum(is.na(db))
# get number of missing values by column
na_counts <- colSums(is.na(db))
```

```{r, echo=FALSE}
attach(db)
names(db)
#is R understanding well my factor variables?
sapply(db,class)

```

## We separate numerical and categorical features

```{r, echo=FALSE}
library(dplyr)
numeriques<-which(sapply(db,is.numeric))
numeriques

categorics<-which(sapply(db,is.factor))
categorics

# Select only the categorical variables
numerical_data <- select_if(db, is.numeric)
categorical_data <- select_if(db, is.factor)

```

## Perform PCA

```{r, echo=FALSE}

# Load the dplyr package
#install.packages("dplyr")
library(dplyr)

# Perform PCA on the data
pca <- prcomp(numerical_data, scale. = TRUE)

# View the results
summary(pca)

#install.packages("corrplot")

```

## Correlation matrix

```{r, echo=FALSE}
# Load the corrplot package
library(corrplot)
# View the correlation matrix only of the numerical values
cor_mat<- cor(numerical_data)
# Set the upper triangle elements to NA
cor_mat[upper.tri(cor_mat)] <- NA

```

## Plot the correlation matrix

```{r, fig.height =16, fig.width = 12, fig.align = "center" }
corrplot(cor_mat, method="circle")
```

```{r}
# pca$sdev
variance<- pca$sdev^2 
# variance
totalVariance<- sum(variance)
# totalVariance
percentageVar<- 100*variance/totalVariance
# percentageVar
barplot(percentageVar)
```


```{r, fig.height =16, fig.width = 12, fig.align = "center" }

# Create a screeplot
barplot(percentageVar, main = "Scree plot", xlab = "Dimensions", ylab = "Percentage of explained variances", col = rainbow(length(percentageVar)))
lines(percentageVar)
points(percentageVar)
text(seq_along(percentageVar), percentageVar, labels = paste0(round(percentageVar,1), "%"), pos=3)
axis(2, at=seq(0,max(percentageVar),5), labels=paste0(seq(0,max(percentageVar),5),'%'))

```

## We use cumulative proportion of variance to easily spot how many components of pca to choose

```{r}
SumPercentageVar <- cumsum(percentageVar)
barplot(SumPercentageVar, main = "Cumulative Scree plot", xlab = "Number of components", ylab = "Cumulative proportion of variance", col = rainbow(length(SumPercentageVar)))
lines(SumPercentageVar)
points(SumPercentageVar)
text(seq_along(SumPercentageVar), SumPercentageVar, labels = paste0(round(SumPercentageVar,1), "%"), pos=3)
axis(2, at=seq(0,max(SumPercentageVar),5), labels=paste0(seq(0,max(SumPercentageVar),5),'%'))
```

In this case we would choose 5

```{r}
# SELECTION OF THE SINGIFICNT DIMENSIONS (keep 80% of total inertia)

nd = 5
pca$rotation # Loadings of the pca (how much of that variable is represented by the principal component)

```

At a quick glance it would seem that the most important variables in terms of the variance be num.persons.worked.for.employer and weeks.worked.in.year since they have the higher absolute value in PC1 which amounts for the most amount of variance.

## We project our data on only 5 dimensions of pca

```{r}
# STORAGE OF THE EIGENVALUES, EIGENVECTORS AND PROJECTIONS IN THE nd DIMENSIONS
# View(pca$x)
dim(pca$x)
dim(numerical_data)
numerical_data[2000,]
pca$x[2000,]

Psi = pca$x[,1:nd]
dim(Psi)
Psi[2000,]

# Observations on significant PCAs.
SignificantCoord = pca$x[,1:nd]
CorrelationSignificant = cor(numerical_data,SignificantCoord) 

```

## We store some data to plot later

```{r}
iden = row.names(numerical_data)
etiq = names(numerical_data)
ze = rep(0,length(etiq)) # WE WILL NEED THIS VECTOR AFTERWARDS FOR THE GRAPHICS

```

# Plot of quality of variables using the square cosine
```{r, fig.height =16, fig.width = 12, fig.align = "center"}
SquareCos = CorrelationSignificant^2
color_palette <- c("#FC8D62", "#FDCC8A", "#FFFFB3", "#A6D854", "#1A9850")
corrplot(SquareCos, is.corr=FALSE, title="Quality of the variables with the PCAs",col =color_palette, bg = "#F1F1F1", tl.col = 'black', mar=c(3,0,2,0))
```

# Individuals Plot

```{r, fig.height =16, fig.width = 12, fig.align = "center"}
color_palette1 <- c("#4B5D67", "#322F3D", "#59405C", "#87556F")
colorIteration <- 1

for (i in seq(nd-1)) {
  for (j in seq(i+1, nd)) {
    plot(SignificantCoord[, i], SignificantCoord[, j],
         col = color_palette1[(colorIteration %% length(color_palette1)) + 1],
         main = "Individuals plot",
         xlab = paste0("PCA", i, " (", percentageVar[i], ' %)'),
         ylab = paste0("PCA", j, " (", percentageVar[j], ' %)'),
         pch = 19,
         cex = 2.5, asp = 0.5)
    abline(h = 0, v = 0, lty = "dashed")
    
    
    colorIteration <- colorIteration + 1
  }
}
```

# Correlation Circle
```{r, fig.height =16, fig.width = 12, fig.align = "center"}

```

```{r}
eje1<-1
#eje2<-3
eje2<-2

plot(Psi[,eje1],Psi[,eje2])
text(Psi[,eje1],Psi[,eje2],labels=iden, cex=0.5)
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")


```

For example let's take a look at 3402 to make sure it makes sense: as we can see PC1 corresponds roughly to 4 and PC2 is in the middle of -40 and -30

```{r}
Psi[3402,]

```

```{r}
#Projection of variables
Phi = cor(numerical_data,Psi)
#View(Phi)

#select your axis

X<-Phi[,eje1]
Y<-Phi[,eje2]
```

View

```{r}

for (i in seq(nd-1)) {  
  for (j in seq(i+1, nd)) {
  
    X<-Phi[,i]
    Y<-Phi[,j]
    plot(Psi[,i],Psi[,j],type="n", ylim = c(-1, 1), xlim = c(-1, 1), xlab = paste("Dim", i), ylab = paste("Dim", j), main = paste("Projection of numeric variables in X: ", i, ", Y: ", j, sep = "") )
    axis(side=1, pos= 0, labels = F)
    axis(side=3, pos= 0, labels = F)
    axis(side=2, pos= 0, labels = F)
    axis(side=4, pos= 0, labels = F)
    arrows(ze, ze, X, Y, length = 0.07,col="blue")
    text(X,Y,labels=etiq,col="darkblue", cex=0.7)


  }
}

```

We can see something that might be interesting here, that is, there is a group of children grouped up at the same place in the following graph.

```{r}
# PROJECTION OF ILLUSTRATIVE qualitative variables on individuals' map
# PROJECCI? OF INDIVIDUALS DIFFERENTIATING THE Dictamen
# (we need a numeric Dictamen to color)


varcat=factor(db[,5])
plot(Psi[,1],Psi[,2],col=varcat)
axis(side=1, pos= 0, labels = F, col="darkgray")
axis(side=3, pos= 0, labels = F, col="darkgray")
axis(side=2, pos= 0, labels = F, col="darkgray")
axis(side=4, pos= 0, labels = F, col="darkgray")
legend("bottomleft",levels(factor(varcat)),pch=1,col=c(1,2), cex=0.6)

```

```{r,  fig.show='hold'}
#categorical_data
# varcat=factor(categorical_data[,6])
# plot(Psi[,1],Psi[,2],col=varcat)
# axis(side=1, pos= 0, labels = F, col="darkgray")
# axis(side=3, pos= 0, labels = F, col="darkgray")
# axis(side=2, pos= 0, labels = F, col="darkgray")
# axis(side=4, pos= 0, labels = F, col="darkgray")
# legend("bottomleft",levels(factor(varcat)),pch=1,col=c(1,2), cex=0.6)
# create an empty list to store the plots
plot_list <- list()

# iterate through all columns of the categorical_data dataset
for (i in 1:ncol(categorical_data)) {
  
  # extract the ith column
  varcat <- factor(categorical_data[,i])
  
  # create a plot
  plot(Psi[,1],Psi[,2],col=varcat)
  axis(side=1, pos= 0, labels = F, col="darkgray")
  axis(side=3, pos= 0, labels = F, col="darkgray")
  axis(side=2, pos= 0, labels = F, col="darkgray")
  axis(side=4, pos= 0, labels = F, col="darkgray")
  legend("bottomleft",levels(factor(varcat)),pch=1,col=c(1,2), cex=0.6)
}
```



```{r}
#select your qualitative variable
k<-5 #dictamen in credsco

varcat<-factor(db[,k])
fdic1 = tapply(Psi[,eje1],varcat,mean)
fdic2 = tapply(Psi[,eje2],varcat,mean) 
plot.new()
points(fdic1,fdic2,pch=16,col="blue", labels=levels(varcat))
text(fdic1,fdic2,labels=levels(varcat),col="black", cex=0.7)

#Now we project both cdgs of levels of a selected qualitative variable without
#representing the individual anymore




```

```{r}
plot(Psi[,eje1],Psi[,eje2],type="n")
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")

```

```{r}
#select your qualitative variable
k<-1 #dictamen in credsco

#varcat<-dd[,k]
#fdic1 = tapply(Psi[,eje1],varcat,mean)
#fdic2 = tapply(Psi[,eje2],varcat,mean) 
plot.new()

#points(fdic1,fdic2,pch=16,col="blue", labels=levels(varcat))
text(fdic1,fdic2,labels=levels(varcat),col="blue", cex=0.7)


#all qualitative together
plot(Psi[,eje1],Psi[,eje2],type="n")
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")


```
