---
title: "PCA"
output: pdf_document
date: "2023-03-16"
---
# Set the working directory
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dirname(dirname(rstudioapi::getActiveDocumentContext()$path)))

library(dplyr)
```

## Read Data

#```{r, echo=FALSE}
# Go to the MD project folder before running the script
#dir <- "C:/Users/polcg/repos/MD" # for some reason the it can't access current directory
# Import preprocessed data
# db <- read.csv(paste0(dir,"/data/Preprocessed_data.csv"),header=T, sep=",", strip.white = TRUE) # take out trailing whitespaces

# read dataframe back in from RDS file
#db <- readRDS(paste0(dir,"/data/Preprocessed_data_rds.rds"))
#```

# Try this to read the data
```{r, echo=FALSE}
# Get the data
db <- readRDS("./data/Preprocessed_data_rds.rds")
```


## We check for missing data

```{r, echo=FALSE}
# Check if there are any missing values, there should not be because they have been imputed, or removed
sum(is.na(db))
# get number of missing values by column
na_counts <- colSums(is.na(db))
```

```{r, echo=FALSE}
attach(db)
names(db)
#is R understanding well my factor variables?
sapply(db,class)

```

## We separate numerical and categorical features

```{r, echo=FALSE}
library(dplyr)
numeriques<-which(sapply(db,is.numeric))
numeriques

categorics<-which(sapply(db,is.factor))
categorics

# Select only the categorical variables
numerical_data <- select_if(db, is.numeric)
categorical_data <- select_if(db, is.factor)

```

## Perform PCA

```{r, echo=FALSE}

# Load the dplyr package
#install.packages("dplyr")
library(dplyr)

# Perform PCA on the data
pca <- prcomp(numerical_data, scale. = TRUE)

# View the results
summary(pca)

#install.packages("corrplot")

```

## Correlation matrix

```{r, echo=FALSE}
# Load the corrplot package
library(corrplot)
# View the correlation matrix only of the numerical values
cor_mat<- cor(numerical_data)
# Set the upper triangle elements to NA
cor_mat[upper.tri(cor_mat)] <- NA

```

## Plot the correlation matrix

```{r, fig.height =16, fig.width = 12, fig.align = "center" }
corrplot(cor_mat, method="circle")
```

```{r}
# pca$sdev
variance<- pca$sdev^2 
# variance
totalVariance<- sum(variance)
# totalVariance
percentageVar<- 100*variance/totalVariance
# percentageVar
barplot(percentageVar)
```


```{r, fig.height =16, fig.width = 12, fig.align = "center" }

# Create a screeplot
barplot(percentageVar, main = "Scree plot", xlab = "Dimensions", ylab = "Percentage of explained variances", col = rainbow(length(percentageVar)))
lines(percentageVar)
points(percentageVar)
text(seq_along(percentageVar), percentageVar, labels = paste0(round(percentageVar,1), "%"), pos=3)
axis(2, at=seq(0,max(percentageVar),5), labels=paste0(seq(0,max(percentageVar),5),'%'))

```

## We use cumulative proportion of variance to easily spot how many components of pca to choose

```{r}
SumPercentageVar <- cumsum(percentageVar)
barplot(SumPercentageVar, main = "Cumulative Scree plot", xlab = "Number of components", ylab = "Cumulative proportion of variance", col = rainbow(length(SumPercentageVar)))
lines(SumPercentageVar)
points(SumPercentageVar)
text(seq_along(SumPercentageVar), SumPercentageVar, labels = paste0(round(SumPercentageVar,1), "%"), pos=3)
axis(2, at=seq(0,max(SumPercentageVar),5), labels=paste0(seq(0,max(SumPercentageVar),5),'%'))
```

In this case we would choose 5

```{r}
# SELECTION OF THE SINGIFICNT DIMENSIONS (keep 80% of total inertia)

nd = 5
pca$rotation # Loadings of the pca (how much of that variable is represented by the principal component)

```

At a quick glance it would seem that the most important variables in terms of the variance be num.persons.worked.for.employer and weeks.worked.in.year since they have the higher absolute value in PC1 which amounts for the most amount of variance.

## We project our data on only 5 dimensions of pca

```{r}
# STORAGE OF THE EIGENVALUES, EIGENVECTORS AND PROJECTIONS IN THE nd DIMENSIONS
# View(pca$x)
dim(pca$x)
dim(numerical_data)
numerical_data[2000,]
pca$x[2000,]

Psi = pca$x[,1:nd]
dim(Psi)
Psi[2000,]

# Observations on significant PCAs.
SignificantCoord = pca$x[,1:nd]
CorrelationSignificant = cor(numerical_data,SignificantCoord) 

```

## We store some data to plot later

```{r}
iden = row.names(numerical_data)
etiq = names(numerical_data)
ze = rep(0,length(etiq)) # WE WILL NEED THIS VECTOR AFTERWARDS FOR THE GRAPHICS

```

# Plot of quality of variables using the square cosine
```{r, fig.height =16, fig.width = 12, fig.align = "center"}
SquareCos = CorrelationSignificant^2
color_palette <- c("#FC8D62", "#FDCC8A", "#FFFFB3", "#A6D854", "#1A9850")
corrplot(SquareCos, is.corr=FALSE, title="Quality of the variables with the PCAs",col =color_palette, bg = "#F1F1F1", tl.col = 'black', mar=c(3,0,2,0))
```

# Individuals Plot

```{r, fig.height =16, fig.width = 12, fig.align = "center"}
color_palette1 <- c("#4B5D67", "#322F3D", "#59405C", "#87556F")
colorIteration <- 1

for (i in seq(nd-1)) {
  for (j in seq(i+1, nd)) {
    plot(SignificantCoord[, i], SignificantCoord[, j],
         col = color_palette1[(colorIteration %% length(color_palette1)) + 1],
         main = "Individuals plot",
         xlab = paste0("PCA", i, " (", percentageVar[i], ' %)'),
         ylab = paste0("PCA", j, " (", percentageVar[j], ' %)'),
         pch = 19,
         cex = 2.5, asp = 0.5)
    abline(h = 0, v = 0, lty = "dashed")
    
    
    colorIteration <- colorIteration + 1
  }
}
```

# Correlation Circle
```{r, fig.height =16, fig.width = 12, fig.align = "center"}

```

```{r}
eje1<-1
#eje2<-3
eje2<-2

plot(Psi[,eje1],Psi[,eje2])
text(Psi[,eje1],Psi[,eje2],labels=iden, cex=0.5)
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")


```

For example let's take a look at 3402 to make sure it makes sense: as we can see PC1 corresponds roughly to 4 and PC2 is in the middle of -40 and -30

```{r}
Psi[3402,]

```

```{r}
#Projection of variables
Phi = cor(numerical_data,Psi)
#View(Phi)

#select your axis

X<-Phi[,eje1]
Y<-Phi[,eje2]
```

View

```{r}

for (i in seq(nd-1)) {  
  for (j in seq(i+1, nd)) {
  
    X<-Phi[,i]
    Y<-Phi[,j]
    plot(Psi[,i],Psi[,j],type="n", ylim = c(-1, 1), xlim = c(-1, 1), xlab = paste("Dim", i), ylab = paste("Dim", j), main = paste("Projection of numeric variables in X: ", i, ", Y: ", j, sep = "") )
    axis(side=1, pos= 0, labels = F)
    axis(side=3, pos= 0, labels = F)
    axis(side=2, pos= 0, labels = F)
    axis(side=4, pos= 0, labels = F)
    arrows(ze, ze, X, Y, length = 0.07,col="blue")
    text(X,Y,labels=etiq,col="darkblue", cex=0.7)

  }
}

```

We can see something that might be interesting here, that is, there is a group of children grouped up at the same place in the following graph.

```{r,  fig.show='hold'}
# iterate through all columns of the categorical_data dataset
for (i in 1:ncol(categorical_data)) {
  # extract the ith column
  varcat <- factor(categorical_data[,i])
  
  # create a plot
  plot(Psi[,1],Psi[,2],col=varcat)
  axis(side=1, pos= 0, labels = F, col="darkgray")
  axis(side=3, pos= 0, labels = F, col="darkgray")
  axis(side=2, pos= 0, labels = F, col="darkgray")
  axis(side=4, pos= 0, labels = F, col="darkgray")
  legend("bottomleft",levels(factor(varcat)),pch=1,col=c(1,2), cex=0.6)
}
```

```{r}
#select your qualitative variable
k<-1 #dictamen in credsco

plot(Psi[,eje1],Psi[,eje2],type="n")
varcat<-db[,k]
fdic1 = tapply(Psi[,eje1],varcat,mean)
fdic2 = tapply(Psi[,eje2],varcat,mean) 

points(fdic1,fdic2,pch=16,col="blue", labels=levels(varcat))
text(fdic1,fdic2,labels=levels(varcat),col="blue", cex=0.7)


#all qualitative together
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")


```
```{r}
#all qualitative together
plot(Psi[,eje1],Psi[,eje2],type="n")
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")
#nominal qualitative variables

dcat<-c(1,3,6:7)
#divide categoricals in several graphs if joint representation saturates

#build a palette with as much colors as qualitative variables 

#colors<-c("blue","red","green","orange","darkgreen")
#alternative
colors<-rainbow(length(dcat))

c<-1
for(k in dcat){
  seguentColor<-colors[c]
fdic1 = tapply(Psi[,eje1],db[,k],mean)
fdic2 = tapply(Psi[,eje2],db[,k],mean) 

text(fdic1,fdic2,labels=levels(factor(db[,k])),col=seguentColor, cex=0.6)
c<-c+1
}
legend("bottomleft",names(db)[dcat],pch=1,col=colors, cex=0.6)
```

```{r}
# ES UN FACTOR D'ESCALA PER DIBUIXAR LES FLETXES MES VISIBLES EN EL GRAFIC
#fm = round(max(abs(Psi[,1]))) 
fm=20

#scale the projected variables
#X<-fm*U[,eje1]
#Y<-fm*U[,eje2]

#represent numerical variables in background
plot(Psi[,eje1],Psi[,eje2],type="n",xlim=c(-1,1), ylim=c(-3,1))
#plot(X,Y,type="none",xlim=c(min(X,0),max(X,0)))
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")

#add projections of numerical variables in background
arrows(ze, ze, X, Y, length = 0.07,col="lightgray")
text(X,Y,labels=etiq,col="gray", cex=0.7)

#add centroids
c<-1
for(k in dcat){
  seguentColor<-colors[c]
  
  fdic1 = tapply(Psi[,eje1],db[,k],mean)
  fdic2 = tapply(Psi[,eje2],db[,k],mean) 
  
  #points(fdic1,fdic2,pch=16,col=seguentColor, labels=levels(dd[,k]))
  text(fdic1,fdic2,labels=levels(factor(db[,k])),col=seguentColor, cex=0.6)
  c<-c+1
}
legend("bottomleft",names(db)[dcat],pch=1,col=colors, cex=0.6)


#add ordinal qualitative variables. Ensure ordering is the correct

dordi<-c(8)


levels(factor(db[,dordi[1]]))
#reorder modalities: when required
db[,dordi[1]] <- factor(db[,dordi[1]], ordered=TRUE,  levels= c("WorkingTypeUnknown","altres sit","temporal","fixe","autonom"))
levels(db[,dordi[1]])

c<-1
col<-1
for(k in dordi){
  seguentColor<-colors[col]
  fdic1 = tapply(Psi[,eje1],db[,k],mean)
  fdic2 = tapply(Psi[,eje2],db[,k],mean) 
  
  #points(fdic1,fdic2,pch=16,col=seguentColor, labels=levels(dd[,k]))
  #connect modalities of qualitative variables
  lines(fdic1,fdic2,pch=16,col=seguentColor)
 text(fdic1,fdic2,labels=levels(db[,k]),col=seguentColor, cex=0.6)
  c<-c+1
  col<-col+1
}
legend("topleft",names(db)[dordi],pch=1,col=colors[1:length(dordi)], cex=0.6)

#using our own colors palette
# search palettes in internet. One might be https://r-charts.com/es/colores/

colors<-c("red", "blue", "darkgreen", "orange", "violet", "magenta", "pink")

#represent numerical variables in background
```

```{r}
#represent numerical variables in background
plot(Psi[,eje1],Psi[,eje2],type="n",xlim=c(-1,1), ylim=c(-3,1))
#plot(X,Y,type="none",xlim=c(min(X,0),max(X,0)))
axis(side=1, pos= 0, labels = F, col="cyan")
axis(side=3, pos= 0, labels = F, col="cyan")
axis(side=2, pos= 0, labels = F, col="cyan")
axis(side=4, pos= 0, labels = F, col="cyan")

#add projections of numerical variables in background
arrows(ze, ze, X, Y, length = 0.07,col="lightgray")
text(X,Y,labels=etiq,col="gray", cex=0.7)

#add centroids
c<-1
for(k in dcat){
  seguentColor<-colors[c]
  
  fdic1 = tapply(Psi[,eje1],db[,k],mean)
  fdic2 = tapply(Psi[,eje2],db[,k],mean) 
  
  #points(fdic1,fdic2,pch=16,col=seguentColor, labels=levels(dd[,k]))
  text(fdic1,fdic2,labels=levels(factor(db[,k])),col=seguentColor, cex=0.6)
  c<-c+1
}
legend("bottomleft",names(db)[dcat],pch=19,col=colors, cex=0.6)


#add ordinal qualitative variables. Ensure ordering is the correct

dordi<-c(8)


levels(factor(db[,dordi[1]]))
#reorder modalities: when required
db[,dordi[1]] <- factor(db[,dordi[1]], ordered=TRUE,  levels= c("WorkingTypeUnknown","altres sit","temporal","fixe","autonom"))
levels(db[,dordi[1]])

c<-1
col<-length(dcat)+1
for(k in dordi){
  seguentColor<-colors[col]
  fdic1 = tapply(Psi[,eje1],db[,k],mean)
  fdic2 = tapply(Psi[,eje2],db[,k],mean) 
  
  #points(fdic1,fdic2,pch=16,col=seguentColor, labels=levels(dd[,k]))
  #connect modalities of qualitative variables
  lines(fdic1,fdic2,pch=16,col=seguentColor)
  text(fdic1,fdic2,labels=levels(db[,k]),col=seguentColor, cex=0.6)
  c<-c+1
  col<-col+1
}
legend("topleft",names(db)[dordi],pch=19,col=colors[col:col+length(dordi)-1], cex=0.6)


#Make two complementary factorial maps

colors<-c("red", "blue", "darkgreen", "orange", "violet", "magenta", "pink")


```
This shows how much the numerical values vary from the mean
```{r}
# PROJECTION OF ILLUSTRATIVE qualitative variables on individuals' map
# PROJECCI? OF INDIVIDUALS DIFFERENTIATING THE Dictamen
# (we need a numeric Dictamen to color)

varcat=factor(db[,1])
plot(Psi[,1],Psi[,2],col=varcat)
axis(side=1, pos= 0, labels = F, col="darkgray")
axis(side=3, pos= 0, labels = F, col="darkgray")
axis(side=2, pos= 0, labels = F, col="darkgray")
axis(side=4, pos= 0, labels = F, col="darkgray")
legend("bottomleft",levels(varcat),pch=1,col=c(1,2), cex=0.6)


# Overproject THE CDG OF  LEVELS OF varcat
fdic1 = tapply(Psi[,1],varcat,mean)
fdic2 = tapply(Psi[,2],varcat,mean) 

text(fdic1,fdic2,labels=levels(factor(varcat)),col="cyan", cex=0.75)

```


