# Go to the MD project folder before running the script
#install.packages("rmarkdown")
#library("rmarkdown")


```{r}
# ENVIROMENT 

dir <- getwd()     # descomenta esto si lo quieres usar en tu ordenador
dir <- "/home/he/Desktop/MD/" 
col_names <- variable_names <- c("age", "class of worker", "detailed industry recode", 
                    "detailed occupation recode", "education", "wage per hour", 
                    "enroll in edu inst last wk", "marital stat", "major industry code", 
                    "major occupation code", "race", "hispanic origin", "sex", 
                    "member of a labor union", "reason for unemployment", 
                    "full or part time employment stat", "capital gains", "capital losses", 
                    "dividends from stocks", "tax filer stat", "region of previous residence", 
                    "state of previous residence", "detailed household and family stat", 
                    "detailed household summary in household", "instance weight", 
                    "migration code-change in msa", "migration code-change in reg", 
                    "migration code-move within reg", "live in this house 1 year ago", 
                    "migration prev res in sunbelt", "num persons worked for employer", 
                    "family members under 18", "country of birth father", 
                    "country of birth mother", "country of birth self", "citizenship", 
                    "own business or self employed", "fill inc questionnaire for veteran's admin", 
                    "veterans benefits", "weeks worked in year", "year")


db <- read.csv(paste0(dir,"/census-income.data"),header=F, sep=",", strip.white = TRUE) # take out trailing whitespaces

# Assign the column names to the data frame
colnames(db) <- col_names

```
#mean(is.na(db)) * 100
#num_question_marks <- sum(db == "?", na.rm = TRUE)
#total_cells = ncol(db) * nrow(db)


```{r}

# transformacion missing values to NA
db[db == "Not in universe"] <- NA
db[db == "Not in universe or children"] <- NA
db[db == "?"] <- NA
db[db == "Not in universe under 1 year old"] <- NA

```


```{r}
#calculo % de missing data para cada columna
colMeans(is.na(db)) * 100

# Eliminar columnas 
db$year <- NULL      # columna que no aporta ninguna informacion
#db <- db[, colMeans(is.na(db)) < 0.9]  # Eliminar todas %missing data > 90%
db$`enroll in edu inst last wk` <- NULL
db$`member of a labor union` <- NULL
db$`reason for unemployment` <- NULL
db$`region of previous residence` <- NULL
db$`state of previous residence` <- NULL
db$`migration prev res in sunbelt` <- NULL
db$`fill inc questionnaire for veteran's admin` <- NULL

# 70% missing data:
db$`family members under 18` <- NULL

# 50% missing data:
db$`migration code-change in msa` <- NULL
db$`migration code-change in reg` <- NULL
db$`migration code-move within reg` <- NULL
db$`live in this house 1 year ago` <- NULL

```

```{r}
# Imputacion: 

# sudo apt-get install libcurl4-openssl-dev
# sudo apt-get install libxml2-dev

# Convert categorical columns to factors for imputation
binaryVars <- c("class of worker","education","marital stat","major industry code","major occupation code","race","hispanic origin","sex","full or part time employment stat","tax filer stat","detailed household and family stat","detailed household summary in household","country of birth father","country of birth mother","country of birth self","citizenship")

#Recorre las columnas señaladas en binaryVars y le hace factor
for(col in binaryVars){
  newCol <- as.character(db[,col])
  newCol <- factor(newCol)
  db[,col] <- newCol
}


# categorical: mode
library(imputeTS)
db$`hispanic origin` <- impute(db$`hispanic origin`, method = "mode")
#db$`hispanic origin` <- na_mode(db$`hispanic origin`, option = "mode")

# categorical: hot-deck, feature is related to other features in your dataset

imputedFa <- impute(db$`country of birth father`, method = "hotdeck")
imputedMo <- impute(db$`country of birth mother`, method = "hotdeck")
imputedSe <- impute(db$`country of birth self`, method = "hotdeck")

db$`country of birth father` <- imputedFa
db$`country of birth mother` <- imputedMo
db$`country of birth self` <- imputedSe

#check if no missing data:
#any(is.na(db$country_father))
#any(is.na(db$country_mother))
#any(is.na(db$country_fself))

# replace NA values in the age column with "Do not know"
# class of worker: 1-> Federal government; 2-> Local government ... -1-> Do not Know
db$`class of worker` <- ifelse(is.na(db$`class of worker`), "-1", db$`class of worker`)
db$`major industry code` <- ifelse(is.na(db$`major industry code`), "-1", db$`major industry code`)
db$`major occupation code` <- ifelse(is.na(db$`major occupation code`), "-1", db$`major occupation code`)



#library(DMwR2)

# Impute missing values using kNN
#db_imputed <- knnImputation(db, k = 5);


#write.csv(db, file = "new_imputed_data.csv", row.names = TRUE)

# Check if there are still missing values
any(is.na(db))

```


```{r}





```

db <- db[, -26]
db <- db[, -27]
db <- db[, -28]
db <- db[, -30]

install.packages("caret")
library(caret)

#define one-hot encoding function
dummy <- dummyVars(" ~ .", data=db)

#perform one-hot encoding on data frame
db_hot <- data.frame(predict(dummy, newdata=db))
# Need to use one-hot encoding first

# Write the data frame to a CSV file
write.table(data, file = "data/processed_data/one_hot_db.csv", sep = ",", row.names = FALSE)

corrplot(cor_mat, method="circle")
colnames(db) <- c("age", "class of worker", "detailed industry recode", "detailed occupation recode", "education", "wage per hour", "enroll in edu inst last wk", "marital stat", "major industry code", "major occupation code", "race", "hispanic origin", "sex", "member of a labor union", "reason for unemployment", "full or part time employment stat", "capital gains", "capital losses", "dividends from stocks", "tax filer stat", "region of previous residence", "state of previous residence", "detailed household and family stat", "detailed household summary in household", "instance weight", "migration code-change in msa", "migration code-change in reg", "migration code-move within reg", "live in this house 1 year ago", "migration prev res in sunbelt", "num persons worked for employer", "family members under 18", "country of birth father", "country of birth mother", "country of birth self", "citizenship", "own business or self employed", "fill inc questionnaire for veteran's admin", "veterans benefits", "weeks worked in year", "year", "income")
